---
title: "Analysis of Wildfires in the U.S. from 1992-2015"
subtitle: "STAT 430 Final Presentation"
author: "Polina Bondarenko, Jason Han, Ben Lipka, Milan Rajababoo"
date: "12/17/2020"
output: 
  ioslides_presentation:
    logo: logo.png
    widescreen: true
    smaller: true
runtime: shiny
---
  
<script src="https://ajax.googleapis.com/ajax/libs/jquery/1.12.2/jquery.min.js"></script>
  
<script>
  $(document).ready(function() {
    $('slide:not(.title-slide, .backdrop, .segue)').append('<footer label=\"University of Illinois at Urbana-Champaign 2020\"></footer>');    
  })
</script>
  
<style>
  footer:after {
    content: attr(label);
    font-size: 12pt;
    position: absolute;
    bottom: 20px;
    left: 100px;
    line-height: 1.9;
  }
}
</style>
 
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

- Collection of the data was funded by the US Department of Agriculture to support the national Fire Program Analysis (FPA) system.
- Wildfire data set contains spatial and temporal information about wildfires which have occurred in the United States between 1992 and 2015. 
- There are 1.88 million total observations taken over the 24 year time period.
- Extracted the data using SQLite
- Subsetted the data by selecting important columns, reformatting the date columns, and adding a column for fire duration (in days). 
- Removed NA values and left subset of 402628 observations and 16 variables.
- Objective: Identify potential trends in wildfire data, visualize discovered results, and create a model in order to predict future wildfires.

## Shiny App

<div style="margin-top:-120px">
  
```{r, include=FALSE}
#make sure working directory is correct
library(shiny)
library(usmap)
library(RSQLite)
library(dbplyr)
library(dplyr)
library(purrr)
library(ggplot2)
library(xts)
library(ggfortify)
library(ggthemes)
library(maps)
library(mapdata)
library(leaflet)
library(mapproj)
library(stats)
library(data.table)
#library(hrbrthemes) #Comment out this + ggiraph
#this package might cause you problems. you might need to download something called XQuartz
library(dygraphs)
library(plotly)
library(highcharter)
#library(ggiraph) #comment out this + hrbrthemes and it should work but not what we want
#this package might cause you problems. you might need to download something called XQuartz
library(shinythemes)
# create db connection
conn <- dbConnect(SQLite(), 'FPA_FOD_20170508.sqlite')
# pull the fires table into RAM
fires <- tbl(conn, "Fires") %>% collect()
# check size
print(object.size(fires), units = 'Gb')
# disconnect from db
# subset by selecting important columns, reformatting date columns, and adding fire duration (in days) column
subsetFire <- dbGetQuery(conn, 
                         "SELECT NWCG_REPORTING_AGENCY, NWCG_REPORTING_UNIT_NAME, FIRE_NAME, FIRE_YEAR, 
                       date(DISCOVERY_DATE) as DISCOVERY_DATE, DISCOVERY_DOY, STAT_CAUSE_CODE, STAT_CAUSE_DESCR, 
                       date(CONT_DATE) as CONT_DATE, CONT_DOY, FIRE_SIZE, FIRE_SIZE_CLASS, OWNER_DESCR, STATE, COUNTY,
                       FIPS_CODE,FIPS_NAME, (CONT_DOY-DISCOVERY_DOY+1) as DURATION_DAYS
                       FROM fires;")
# remove NA values: over 1.4 million observations removed
#Dataset we will use for the rest of the code
cleanedFire <- na.omit(subsetFire)
cleanedFire <- cleanedFire %>% filter(DURATION_DAYS>0)
cleanedFire$fips <- paste0(fips(cleanedFire$STATE),cleanedFire$FIPS_CODE)
dbDisconnect(conn)
```

```{r, echo=FALSE}
#### data related to maps
options(dplyr.summarise.inform = FALSE)
fires_state <- cleanedFire %>% 
  select(state=STATE, DURATION_DAYS, FIRE_SIZE) %>%
  group_by(state) %>%
  summarize(n = n(), DURATION_DAYS=mean(DURATION_DAYS), FIRE_SIZE=mean(FIRE_SIZE))
fires_county <- cleanedFire %>% 
  select(fips, DURATION_DAYS, FIRE_SIZE) %>%
  group_by(fips) %>%
  summarize(n = n(), DURATION_DAYS=mean(DURATION_DAYS), FIRE_SIZE=mean(FIRE_SIZE))
#### data related to bar graphs
fires_state_yr <- cleanedFire %>% 
  select(state=STATE, FIRE_YEAR) %>%
  group_by(FIRE_YEAR,state) %>%
  summarize(number_of_fires = n())
### code pertaining to pie charts
causes_count_df = cleanedFire %>%
  count(STAT_CAUSE_DESCR)
states_count_df = cleanedFire%>%
  count(STATE)
owner_count_df = cleanedFire %>%
  count(OWNER_DESCR)
firesize_count_df = cleanedFire%>%
  count(FIRE_SIZE_CLASS)
#### SHINY
ui <- shinyUI(navbarPage(theme = shinytheme("united"), "Wildfires in the United States",
                         
                         tabPanel("Map Visualization",
                                  sidebarPanel(
                                    # Input: Selector for variable to plot ----
                                    selectInput("variable", "Variable:",
                                                c("Total Count" = "n",
                                                  "Wildfire Duration" = "DURATION_DAYS",
                                                  "Wildfire Size" = "FIRE_SIZE")), 
                                  ),
                                  
                                  # Main panel for displaying outputs ----
                                  mainPanel(
                                    # Output: Plot of the requested variable ----
                                    plotOutput("usFirePlot"),
                                  )
                         ),
                         
                         tabPanel("State Map Visualization",
                                  sidebarPanel(
                                    # Input: Selector for variable to plot ----
                                    selectInput("variable2", "Variable:",
                                                c("Total Count" = "n",
                                                  "Wildfire Duration" = "DURATION_DAYS",
                                                  "Wildfire Size" = "FIRE_SIZE")), 
                                    selectInput("state", "State:",
                                                c(state.abb)),
                                  ),
                                  
                                  # Main panel for displaying outputs ----
                                  mainPanel(
                                    # Output: Plot of the requested variable ----
                                    plotOutput("stateFirePlot")
                                  )
                         ),
                         
                         tabPanel("Wildfires by Year",
                                  sidebarPanel(
                                    # Input: Selector for choosing dataset ----
                                    selectInput(inputId = "dataset",
                                                label = "Choose a State:",
                                                choices = c(state.abb)),
                                  ),
                                 mainPanel("Bar Plots",
                                          fluidRow(
                                            splitLayout(cellWidths = c("50%", "50%"), plotOutput("plot"), plotOutput("overallFires"))
                                          ),
                                          tableOutput("summary"),
                                )
                         ),
                         tabPanel("Pie Chart Visualization",
                                  sidebarPanel(
                                    # Input: Selector for variable to plot ----
                                    selectInput(inputId = "pie",
                                                label = "Choose a Pie Chart:",
                                                choices = c("Fire Cause", "Fire Size", "Owner of Land")),
                                  ),
                                  # Main panel for displaying outputs ----
                                  
                                  mainPanel(
                                    # Output: Plot of the requested variable ----
                                    tabsetPanel(type = "tabs",
                                                tabPanel("Pie Chart", highchartOutput("Pieplot"))
                                    )
                                  )
                         )
                    
)
)
# Define server logic to summarize and view selected dataset ----
server <- function(input, output) {
  
  # Make title for plot based on input variable ----
  name <- reactive({
    if (input$variable=="n") {
      paste("Total Number of Wildfires")
    } else if (input$variable=="DURATION_DAYS") {
      paste("Average Wildfire Duration")
    } else if (input$variable=="FIRE_SIZE") {
      paste("Average Size of Wildfires")
    }
  })
  
  # Make caption for plot based on input variable ----
  caption <- reactive({
    if (input$variable=="n") {
      paste("Number of Wildfires")
    } else if (input$variable=="DURATION_DAYS") {
      paste("Wildfire Duration (Days)")
    }  else if (input$variable=="FIRE_SIZE") {
      paste("Wildfire Size (Sq. Acres)")
    }
  })
  
  # Function to plot the requested variable ----
  plotUSFunc <- function() {
    plot_usmap(data = fires_state, values = input$variable, color = "gray", labels=TRUE) + 
      scale_fill_continuous(name = caption(), label = scales::comma, type = "viridis") + 
      labs(title = paste0(name()," by State between 1992 and 2015"), 
           caption = "Source: Forest Service Research Data Archive") +
      theme(plot.title = element_text(face="bold", size = 12, hjust = 0.5), plot.caption = element_text(size = 12, hjust=0), 
            legend.position = "right", legend.title=element_text(size=12), legend.text = element_text(size=11), 
            plot.margin = margin(.1,.1,.1,.1, "cm"), panel.background = element_rect(fill = "lightblue"))
  }
  
  name2 <- reactive({
    if (input$variable2=="n") {
      paste("Total Number of Wildfires")
    } else if (input$variable2=="DURATION_DAYS") {
      paste("Average Wildfire Duration")
    } else if (input$variable2=="FIRE_SIZE") {
      paste("Average Size of Wildfires")
    }
  })
  
  # Make caption for plot based on input variable ----
  caption2 <- reactive({
    if (input$variable2=="n") {
      paste("Number of Wildfires")
    } else if (input$variable2=="DURATION_DAYS") {
      paste("Wildfire Duration (Days)")
    }  else if (input$variable2=="FIRE_SIZE") {
      paste("Wildfire Size (Sq. Acres)")
    }
  })
  
  plotStateFunc <- function() {
    plot_usmap(regions = "county", include=c(input$state), data = fires_county, values = input$variable2, color = "gray") + 
      scale_fill_continuous(name = caption2(), label = scales::comma, type = "viridis") + 
      labs(title = paste0(name2(), " in ", input$state, " between 1992 and 2015"), caption = "Source: Forest Service Research Data Archive") +
      theme(plot.title = element_text(face="bold", size = 12, hjust = 0.5), plot.caption = element_text(size = 12, hjust=0), 
            legend.position = "right", legend.title=element_text(size=12), legend.text = element_text(size=11), 
            plot.margin = margin(.1,.1,.1,.1, "cm"), panel.background = element_rect(fill = "lightblue"))
  }
  
  
  # Generate a plot of the requested variable ----
  output$usFirePlot <- renderPlot({
    plotUSFunc()
  })
  
  output$stateFirePlot <- renderPlot({
    plotStateFunc()
  })
  
  # Return the requested dataset ----
  datasetInput <- reactive({
    df <- fires_state_yr[fires_state_yr$state==input$dataset,]
    df
  })
  
    plotOverall <- function() {
    ggplot(data = fires_state_yr, aes(x=FIRE_YEAR,y = number_of_fires )) +geom_bar(stat = 'identity', fill = 'red') +
      geom_smooth(method = 'lm', se = FALSE, size = 0.4, color = 'black') +
      labs(x = 'Fire Year', y = 'Number of wildfires', title = 'Wildfires by Year in the Entire Country')
  }
  output$overallFires <- renderPlot({
    plotOverall()
  })
  
  
  output$plot <- renderPlot({
    ggplot(data = datasetInput(), aes(x=FIRE_YEAR,y = number_of_fires )) +geom_bar(stat = 'identity', fill = 'red') +
      geom_smooth(method = 'lm', se = FALSE, size = 0.4, color = 'black') +
      labs(x = 'Fire Year', y = 'Number of wildfires', title = 'Wildfires by Year')
  })
  
  # Generate a summary of the data ----
  output$summary <- renderTable({
    summary(datasetInput()[c("number_of_fires")])
  })
  
  # Return the requested pie plot ----
  pieInput <- reactive({
    if (input$pie=="Fire Cause") {
      pie_df <- data.frame(
        count = causes_count_df$n,
        label = causes_count_df$STAT_CAUSE_DESCR
      )
    } else if (input$pie=="Fire Size") {
      pie_df = data.frame(
        count = firesize_count_df$n,
        label = firesize_count_df$FIRE_SIZE_CLASS
      )
    } else if (input$pie=="Owner of Land") {
      pie_df = data.frame(
        count = owner_count_df$n,
        label = owner_count_df$OWNER_DESCR
      ) 
    }
    pie_df
  })
  
  # Function to plot pie ----
  plotPie <- function() {
    pieInput() %>%
      hchart(
        "pie", hcaes(x = label, y = count),
        name = "Wildfire Count"
      )
  }
  
  # Generate a plot of the requested variable ----
  output$Pieplot <- renderHighchart({
    plotPie()
  })
  
}
# Create Shiny app ----
shinyApp(ui = ui, server = server)

```

</div>

## Modeling

- Initial attempt with simple linear regression performed extremely poorly, with little variable correlation
- Decision tree performed much better, but still room for improvement
- After broadening classification groups and refining variables used, decision tree and random forest models performed similarly and reached average of 28-30% accuracy level
  
## Future Investigations 
  
- Further investigate NA values in data
- Utilize other datasets to enhance available wildfire statistics
- Strengthen modeling efforts
  
## Conclusion 
  
- Week 2: project documentation via GitHub.
- Week 3: Rmarkdown for final report and presentation slides.
- Week 4: RSQLite for extracting, cleaning, and subsetting data.
- Week 7: data.table and dplyr libraries for data wrangling and manipulation.
- Week 9: data visualization using packages such as ggplot2.
- Week 10: building a Shiny dashboard.

## References 

- Short, Karen C. 2017. Spatial wildfire occurrence data for the United States, 1992-2015 [FPA_FOD_20170508]. 4th Edition. Fort Collins, CO: Forest Service Research Data Archive. https://doi.org/10.2737/RDS-2013-0009.4
- Walters, Troy. 2017. Wildfire Exploratory Analysis. Version 39. Kaggle. https://www.kaggle.com/captcalculator/wildfire-exploratory-analysis/
