---
title: "Visualizations"
output: html_document
---
** Loading all necessary libraries **
```{r setup, include=FALSE}
library(RSQLite)
library(dbplyr)
library(dplyr)
library(purrr)
library(ggplot2)
library(xts)
library(ggfortify)
library(ggthemes)
library(maps)
library(mapdata)
library(leaflet)
library(mapproj)
library(stats)
library(data.table)
library(hrbrthemes)
library(dygraphs)
library(plotly)
library(highcharter)
library(ggiraph)
```

This setups the fire data file.

```{r}
# create db connection
conn <- dbConnect(SQLite(), 'FPA_FOD_20170508.sqlite')

# pull the fires table into RAM
fires <- tbl(conn, "Fires") %>% collect()
fires_no_lim <- dbGetQuery(conn, 
                           "SELECT NWCG_REPORTING_AGENCY, NWCG_REPORTING_UNIT_NAME, FIRE_NAME, FIRE_YEAR, 
                       date(DISCOVERY_DATE) as DISCOVERY_DATE, DISCOVERY_DOY, STAT_CAUSE_CODE, STAT_CAUSE_DESCR, 
                       date(CONT_DATE) as CONT_DATE, CONT_DOY, FIRE_SIZE, FIRE_SIZE_CLASS, OWNER_DESCR, STATE, COUNTY, 
                       (CONT_DOY-DISCOVERY_DOY+1) as DURATION_DAYS
                       FROM fires 
                       ")

# disconnect from db
dbDisconnect(conn)
```

This next chunk creates the main data tables that I will use to subset in subsequent sections.
```{r}
fires_dt = setDT(fires_no_lim)
fires_dt_omit = na.omit(fires_dt)

```

## PIE CHARTS

#### Data Frames that contain the count and the names

```{r}
causes_count_df = fires_dt_omit %>%
  count(STAT_CAUSE_DESCR)
states_count_df = fires_dt_omit%>%
  count(STATE)
owner_count_df = fires_dt_omit %>%
  count(OWNER_DESCR)
firesize_count_df = fires_dt_omit%>%
  count(FIRE_SIZE_CLASS)
```
#### Data Frames that are going to be used for the pie chart. They use the columns from the data frames above.

```{r}

causes_pie_df <- data.frame(
  count = causes_count_df$n,
  label = causes_count_df$STAT_CAUSE_DESCR
)

states_pie_df = data.frame(
  count = states_count_df$n,
  label = states_count_df$STATE
)

owner_pie_df = data.frame(
  count = owner_count_df$n,
  label = owner_count_df$OWNER_DESCR
)

firesize_pie_df = data.frame(
  count = firesize_count_df$n,
  label = firesize_count_df$FIRE_SIZE_CLASS
)

```
#### I use the High Charter package to create an interactable pie chart. The first chart shows how many times a fire is created by a specific cause. The second chart shows how many times a fire occured in a state. The third chart shows who owns the majoriry of forests that fires. The 4th chart shows the ditribution of the fire size, which are classfied with letters.

```{r}

causes_pie_chart <- causes_pie_df %>%
  hchart(
    "pie", hcaes(x = label, y = count),
    name = "Fire Cause"
  )
causes_pie_chart

states_pie_chart <- states_pie_df %>%
  hchart(
    "pie", hcaes(x = label, y = count),
    name = "State Count"
  )
states_pie_chart

owner_pie_chart <- owner_pie_df %>%
  hchart(
    "pie", hcaes(x = label, y = count),
    name = "Owner Count"
  )
owner_pie_chart


firesize_pie_chart <- firesize_pie_df %>%
  hchart(
    "pie", hcaes(x = label, y = count),
    name = "Owner Count"
  )

firesize_pie_chart
```

## FIRE TIME ANALYSIS

```{r}
## Check for irregularities in days variable

firesx = fires_dt[DURATION_DAYS > 0]
sum(table(firesx$DURATION_DAYS[ which(firesx$DURATION_DAYS < 0)]))

```

```{r}
## Visualizations

### General fire duration by year

ggplotly(
  ggplot(firesx) + 
  geom_bar(aes(x=FIRE_YEAR, y=DURATION_DAYS, fill=FIRE_SIZE_CLASS), 
           stat = "summary", fun = "mean") + 
  labs(title="Mean Fire Duration by Year", x="Year", y="Mean Fire Duration (Days)") +
  guides(fill=guide_legend(title="Fire Size Classification Code")) +
  theme_light()
)

### Fire duration by state (alaska, nj, washington)

ggplotly(
  ggplot(firesx) + 
  geom_bar(aes(x=STATE, y=DURATION_DAYS), 
           stat = "summary", fun = "mean") + 
  labs(title="Mean Fire Duration by State",x="State",y="Mean Fire Duration (Days)") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
)

### Fire duration by cause

ggplotly(
  ggplot(firesx) + 
  geom_bar(aes(x=STAT_CAUSE_DESCR, y=DURATION_DAYS, fill=FIRE_SIZE_CLASS), 
           stat="summary", fun="mean") + 
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  labs(title="Mean Fire Duration by Cause",x="Cause of Fire",y="Mean Fire Duration (Days)") +
  guides(fill=guide_legend(title="Fire Size Classification Code"))
)

### Fire duration by fire size classification (don't need to show, this is obvious)

ggplotly(
  ggplot(firesx) +
  geom_bar(aes(x=FIRE_SIZE_CLASS, y=DURATION_DAYS), 
           stat="summary", fun="mean")
)

### Fire duration by owner description (entity responsible for managing land at point of origin)
### fish and wildlife service, national park service, us forest service
### likely dealing with larger land?

ggplotly(
  ggplot(firesx) + 
  geom_bar(aes(x=OWNER_DESCR, y=DURATION_DAYS, fill=FIRE_SIZE_CLASS), 
           stat="summary", fun="mean") + 
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) + 
  labs(title="Mean Fire Duration by Owner of Land",x="Owner of Land",y="Mean Fire Duration (Days)")
)


### Fire duration by county in California
#### maybe show only outliers ???

ggplotly(
  ggplot(firesx[ which(firesx$STATE=="CA" & is.na(as.numeric(firesx$COUNTY))),]) + 
  geom_bar(aes(x=COUNTY, y=DURATION_DAYS), 
           stat="summary", fun="mean") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
)
```

